---
title: "LoPresti2019_DataAnalysis"
output:
  pdf_document: default
  html_document:
    df_print: paged
    toc: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 15, fig.height = 5, warning=FALSE, message=FALSE,
                      root.dir = "~/Projects/IBS_Meta-analysis_16S/analysis-individual/LoPresti-2019")
```



############
## IMPORT ##
############

### 1. Librairies

```{r library-import}
library(phyloseq) # for phyloseq object
library(ggplot2)
library(cowplot)
library(tidyverse)
library("plotly") # plot 3D
library("microbiome") # for centered log-ratio
library("coda") # Aitchison distance
library("coda.base") # Aitchison distance
library("vegan") # NMDS
library(pheatmap) # for heatmap
```

### 2. Data

```{r import-data}
# Set path
path <- "~/Projects/IBS_Meta-analysis_16S"

# Import phyloseq object
physeq.lopresti <- readRDS(file.path(path, "phyloseq-objects/physeq_lopresti.rds"))

# Sanity check
physeq.lopresti
```


Phylogenetic tree was computed with the package `phangorn`, and the script was run on a cluster. Let's check we have correctly generated a phylogenetic tree.

```{r plot-tree, fig.height = 6, fig.width = 10}
# Look at the tree
plot_tree(physeq.lopresti, color = "host_disease", ladderize="left")
```



```{r import-for-html, echo = FALSE}
#__________________________________________________________________________
#____________ THIS IS ONLY FOR (quicker) PDF/HTML OUTPUT __________________
#__________________________________________________________________________

# Import phyloseq objects for computing distances
physeq.rel <- readRDS(file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_relative.rds"))
physeq.clr <- readRDS(file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_clr.rds"))
physeq.CSN <- readRDS(file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_CSN.rds"))
physeq.NZcomp <- readRDS(file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_NZcomp.rds"))
```



##################
## DEMOGRAPHICS ##
##################

This dataset has several covariates (gender, age, bmi, sample_storage_duration). We will check whether there is the same distribution of these covariates between healthy and IBS patients.

```{r demographics-stats}
# Number of individuals in each group (keeping just 1 sample per individual)
metadata <- data.frame(sample_data(physeq.lopresti)) %>%
  select(host_disease, host_age, host_sex, host_ID) %>%
  group_by(host_ID) %>%
  summarise_all(first)
  
metadata %>%
  count(host_disease)

# Age
metadata %>%
  group_by(host_disease) %>%
  summarize(mean_age=mean(host_age), sd_age=sd(host_age))
wilcox.test(metadata[metadata$host_disease == "IBS", ]$host_age,
            metadata[metadata$host_disease == "Healthy", ]$host_age) # p=0.25

# Gender
metadata %>%
  count(host_disease, host_sex)
chisq.test(data.frame("Female" = c(12,17),
                      "Male" = c(18,6))) # p=0.02 -> more females than males in IBS group
```




################
## ABUNDANCES ##
################

### 1. Absolute abundances

```{r plot-absolute-abundance, fig.width=10, fig.height = 5}
# Plot Phylum
plot_bar(physeq.lopresti, fill = "Phylum") + facet_wrap("sample_type", scales="free_x") +
  theme(axis.text.x = element_text(size = 5))+
  labs(x = "Samples", y = "Absolute abundance", title = "Lo Presti dataset (2019)") +
  ylim(0,4000)

# Plot Class
plot_bar(physeq.lopresti, fill = "Class")+ facet_wrap("sample_type", scales="free_x") +
  theme(axis.text.x = element_text(size = 5))+
  labs(x = "Samples", y = "Absolute abundance", title = "Lo Presti dataset (2019)")+
  ylim(0,4000)
```

**Sequencing depth** characteristics of the Hugerth dataset:  
- minimum of `r min(sample_sums(physeq.lopresti))` total count per sample  
- median: `r median(sample_sums(physeq.lopresti))` total count per sample  
- maximum of `r max(sample_sums(physeq.lopresti))` total count per sample  



### 2. Relative abundances

```{r plot-relative-abundances, fig.width=10, fig.height = 5, message=FALSE}
# Agglomerate to phylum & class levels
phylum.table <- physeq.lopresti %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt()                                             # Melt to long format

class.table <- physeq.lopresti %>%
  tax_glom(taxrank = "Class") %>%
  transform_sample_counts(function(x) {x/sum(x)} ) %>%
  psmelt()


# Plot relative abundances
ggplot(phylum.table, aes(x = reorder(Sample, Sample, function(x) mean(phylum.table[Sample == x & Phylum == 'Bacteroidota', 'Abundance'])),
                         y = Abundance, fill = Phylum))+
  facet_wrap(~ sample_type + host_disease, scales = "free") + # scales = "free" removes empty lines
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank())+
  labs(x = "Samples", y = "Relative abundance", title = "Lo Presti dataset (2019)")

ggplot(class.table, aes(x = reorder(Sample, Sample, function(x) mean(class.table[Sample == x & Phylum == 'Bacteroidota', 'Abundance'])),
                        y = Abundance, fill = Class))+
  facet_wrap(~ sample_type + host_disease, scales = "free") +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank(),
        legend.key.size = unit(0.2, 'cm'),
        legend.text = element_text(size=8))+
  labs(x = "Samples", y = "Relative abundance", title = "Lo Presti dataset (2019)")
```




### 3. Firmicutes/Bacteroidota ratio

```{r plot-log-ratio, fig.height = 5, fig.width=5}
# Extract abundance of only Bacteroidota and Firmicutes
relevant.covariates <- c('Sample', 'Abundance', 'host_disease', 'host_subtype', 'Phylum', 'sample_type', 'host_ID', 'host_age', 'host_sex')

bacter <- phylum.table %>%
  filter(Phylum == "Bacteroidota") %>%
  select(all_of(relevant.covariates)) %>%
  dplyr::rename(Bacteroidota = Abundance) %>%
  select(-Phylum)

firmi <- phylum.table %>%
  filter(Phylum == "Firmicutes") %>%
  select(all_of(relevant.covariates)) %>%
  dplyr::rename(Firmicutes = Abundance) %>%
  select(-Phylum)

# Calculate log2 ratio Firmicutes/Bacteroidota
ratio.FB <- left_join(x=bacter, y=firmi, by=c('Sample', 'host_disease', 'host_subtype', 'sample_type', 'host_ID', 'host_age', 'host_sex')) %>%
  relocate(Firmicutes, .after=Bacteroidota) %>%
  # Compute log ratios
  mutate(logRatioFB = log2(Firmicutes/Bacteroidota))


# Plot log2 ratio Firmicutes/Bacteroidota
ggplot(ratio.FB, aes(x = host_disease, y = logRatioFB))+
  geom_violin(aes(fill=host_disease))+
  scale_fill_manual(values=scales::alpha(c("blue", "red"), .3))+
  geom_jitter(width=0.1)+
  facet_wrap(~sample_type) +
  labs(x = "",  y = 'Log2(Firmicutes/Bacteroidota)', title = "Firmicutes:Bacteroidota ratio")+
  theme_cowplot()+
  theme(legend.position="none")

# Statistical test sigmoid samples
wilcox.test(ratio.FB[ratio.FB$sample_type == "sigmoid" & ratio.FB$host_disease == "IBS","logRatioFB"],
            ratio.FB[ratio.FB$sample_type == "sigmoid" & ratio.FB$host_disease == "Healthy","logRatioFB"]) # p = 0.4
# Statistical test stool samples
wilcox.test(ratio.FB[ratio.FB$sample_type == "stool" & ratio.FB$host_disease == "IBS","logRatioFB"],
            ratio.FB[ratio.FB$sample_type == "stool" & ratio.FB$host_disease == "Healthy","logRatioFB"]) # p=0.01 significant difference in stool samples


# Plot log2 ratio Firmicutes/Bacteroidota per IBS subtype
ggplot(ratio.FB %>% filter(sample_type == "stool"), aes(x = host_subtype, y = logRatioFB))+
  geom_violin(aes(fill=host_subtype))+
  scale_fill_manual(values=scales::alpha(c("blue", "brown", "red", "orange"), .3))+
  geom_jitter(width=0.1)+
  labs(x = "",  y = 'Log2(Firmicutes/Bacteroidota)', title = "Fecal samples")+
  theme_cowplot()+
  theme(legend.position="none")


# Paired data
ggplot(ratio.FB, aes(x = sample_type, y = logRatioFB))+
  geom_violin(aes(fill=host_disease))+
  scale_fill_manual(values=scales::alpha(c("blue", "red"), .3))+
  geom_point()+
  geom_line(aes(group=host_ID), lwd=0.1) +
  facet_wrap(~host_disease) +
  labs(x = "",  y = 'Log2(Firmicutes/Bacteroidota)', title = "Paired data") +
  theme_cowplot()+
  theme(legend.position="none")

# Statistical test sigmoid vs stool samples
wilcox.test(ratio.FB[ratio.FB$sample_type == "stool" & ratio.FB$host_disease == "Healthy","logRatioFB"],
            ratio.FB[ratio.FB$sample_type == "sigmoid" & ratio.FB$host_disease == "Healthy","logRatioFB"]) # p=0.001
wilcox.test(ratio.FB[ratio.FB$sample_type == "stool" & ratio.FB$host_disease == "IBS","logRatioFB"],
            ratio.FB[ratio.FB$sample_type == "sigmoid" & ratio.FB$host_disease == "IBS","logRatioFB"]) # p=0.04
```




```{r export-plots-1, eval = FALSE, echo = FALSE}
# Save path to plots
path.plots <- file.path(path, "data/analysis-individual/PLOTS/plots-LoPresti-EDA")

# Absolute abundance phylum
jpeg(file.path(path.plots, "absAbundance_phylum.jpg"), width = 4000, height = 2000)
plot_bar(physeq.lopresti, fill = "Phylum") + facet_wrap("sample_type", scales="free_x") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 30),
        legend.text = element_text(size = 50),
        axis.title = element_text(size = 50),
        title = element_text(size = 70),
        strip.text = element_text(size = 50))+
  labs(x = "Samples", y = "Total read count", title = "Lo Presti dataset (2019)")+
  ylim(0, 4000)
dev.off()

# Relative abundances
jpeg(file.path(path.plots, "relAbundance_phylum.jpg"), width = 4000, height = 2000)
ggplot(phylum.table, aes(x = reorder(Sample, Sample, function(x) mean(phylum.table[Sample == x & Phylum == 'Bacteroidota', 'Abundance'])),
                         y = Abundance, fill = Phylum))+
  facet_wrap(~ sample_type + host_disease, scales = "free") + # scales = "free" removes empty lines
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 30),
        legend.text = element_text(size = 50),
        axis.title = element_text(size = 50),
        title = element_text(size = 70),
        strip.text = element_text(size = 50))+
  labs(x = "Samples", y = "Relative abundance", title = "Lo Presti dataset (2019)")
dev.off()

jpeg(file.path(path.plots, "relAbundance_class.jpg"), width = 4000, height = 2000)
ggplot(class.table, aes(x = reorder(Sample, Sample, function(x) mean(phylum.table[Sample == x & Phylum == 'Bacteroidota', 'Abundance'])),
                        y = Abundance, fill = Class))+
  facet_wrap(~ sample_type + host_disease, scales = "free") +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 30),
        legend.text = element_text(size = 50),
        axis.title = element_text(size = 50),
        title = element_text(size = 70),
        strip.text = element_text(size = 50))+
  labs(x = "Samples", y = "Relative abundance", title = "Lo Presti dataset (2019)")
dev.off()


# Firmicutes:Bacteroidota log ratio (healthy/IBS)
jpeg(file.path(path.plots, "ratioFB.jpg"), width = 1500, height = 1500)
ggplot(ratio.FB, aes(x = host_disease, y = logRatioFB))+
  geom_violin(lwd=3, aes(fill=host_disease))+
  geom_jitter(size = 10, width=0.1)+
  scale_fill_manual(values=scales::alpha(c("blue", "red"), .3))+
  facet_wrap(~sample_type) +
  labs(x = "",  y = 'Log2(Firmicutes/Bacteroidota)', title = "Firmicutes:Bacteroidota ratio")+
  theme_cowplot()+
  theme(axis.text = element_text(size = 50),
        strip.text = element_text(size=50),
        legend.position = "none",
        axis.title = element_text(size = 50),
        title = element_text(size = 50))
dev.off()

# Firmicutes:Bacteroidota log ratio (paired data stool/sigmoid)
jpeg(file.path(path.plots, "ratioFB_paired-data.jpg"), width = 1500, height = 1500)
ggplot(ratio.FB, aes(x = sample_type, y = logRatioFB))+
  geom_violin(lwd=3, aes(fill=host_disease))+
  geom_point(size = 10)+
  geom_line(aes(group=host_ID), lwd=0.5) +
  scale_fill_manual(values=scales::alpha(c("blue", "red"), .3))+
  facet_wrap(~host_disease) +
  labs(x = "",  y = 'Log2(Firmicutes/Bacteroidota)', title = "Paired data") +
  theme_cowplot()+
  theme(axis.text = element_text(size = 50),
        strip.text = element_text(size=50),
        legend.position = "none",
        axis.title = element_text(size = 50),
        title = element_text(size = 50))
dev.off()
```




####################
## NORMALIZE DATA ##
####################

```{r normalize-data, eval = FALSE, echo = TRUE}
# Sanity check no sample with less than 500 total count
table(sample_sums(physeq.lopresti)<500) # all FALSE

#____________________________________________________________________
# PHYLOSEQ OBJECT WITH NON-ZERO COMPOSITIONS
physeq.NZcomp <- physeq.lopresti
otu_table(physeq.NZcomp)[otu_table(physeq.NZcomp) == 0] <- 0.5 # pseudocounts

# Sanity check that 0 values have been replaced
# otu_table(physeq.lopresti)[1:5,1:5]
# otu_table(physeq.NZcomp)[1:5,1:5]

# transform into compositions
physeq.NZcomp <- transform_sample_counts(physeq.NZcomp, function(x) x / sum(x) )
table(rowSums(otu_table(physeq.NZcomp))) # check if there is any row not summing to 1

# Save object
saveRDS(physeq.NZcomp, file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_NZcomp.rds"))

#____________________________________________________________________
# PHYLOSEQ OBJECT WITH RELATIVE COUNT (BETWEEN 0 AND 1)
physeq.rel <- physeq.lopresti
physeq.rel <- transform_sample_counts(physeq.rel, function(x) x / sum(x) )

# check the counts are all relative
# otu_table(physeq.lopresti)[1:5, 1:5]
# otu_table(physeq.rel)[1:5, 1:5]

# sanity check
table(rowSums(otu_table(physeq.rel))) # check if there is any row not summing to 1

# save the physeq.rel object
saveRDS(physeq.rel, file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_relative.rds"))

#____________________________________________________________________
# PHYLOSEQ OBJECT WITH COMMON-SCALE NORMALIZATION
physeq.CSN <- physeq.lopresti
physeq.CSN <- transform_sample_counts(physeq.CSN, function(x) (x*min(sample_sums(physeq.CSN))) / sum(x) )

# sanity check
table(rowSums(otu_table(physeq.CSN))) # check that all rows are summing to the same total

# save the physeq.CSN object
saveRDS(physeq.CSN, file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_CSN.rds"))


#____________________________________________________________________
# PHYLOSEQ OBJECT WITH CENTERED LOG RATIO COUNT
physeq.clr <- physeq.lopresti
physeq.clr <- microbiome::transform(physeq.lopresti, "clr") # the function adds pseudocounts itself

# Compare the otu tables in the original phyloseq object and the new one after CLR transformation
# otu_table(physeq.lopresti)[1:5, 1:5] # should contain absolute counts
# otu_table(physeq.clr)[1:5, 1:5] # should all be relative

# save the physeq.rel object
saveRDS(physeq.clr, file.path(path, "data/analysis-individual/LoPresti-2019/02_EDA-LoPresti/physeq_clr.rds"))
```






#######################
## COMPUTE DISTANCES ##
#######################

### 1. UniFrac, Aitchison, Bray-Curtis and Canberra

First, let's look at these four distances of interest.

```{r dist-functions, fig.width = 15, fig.height = 4}
#____________________________________________________________________________________
# Measure distances
getDistances <- function(relPhyseq=physeq.rel, clrPhyseq=physeq.clr, csnPhyseq=physeq.CSN, nzcompPhyseq=physeq.NZcomp){
  # sanity check
  cat("nb samples relPhyseq:", nsamples(relPhyseq), "\n")
  cat("nb samples clrPhyseq:", nsamples(clrPhyseq), "\n")
  cat("nb samples csnPhyseq:", nsamples(csnPhyseq), "\n")
  cat("nb samples nzcompPhyseq:", nsamples(nzcompPhyseq), "\n")
  
  # Compute distances
  print("Unifrac...")
  set.seed(123) # for unifrac, need to set a seed
  glom.UniF <- UniFrac(relPhyseq, weighted=TRUE, normalized=TRUE) # weighted unifrac
  print("Aitchison...")
  glom.ait <- phyloseq::distance(clrPhyseq, method = 'euclidean') # aitchison
  print("Bray des bois...")
  glom.bray <- phyloseq::distance(csnPhyseq, method = "bray") # bray-curtis
  print("Canberra <3...")
  glom.can <- phyloseq::distance(nzcompPhyseq, method = "canberra") # canberra
  
  dist.list <- list("UniF" = glom.UniF, "Ait" = glom.ait, "Canb" = glom.can, "Bray" = glom.bray)
  
  return(dist.list)
}


#____________________________________________________________________________________
# Plot in 2D the distances
plotDistances2D <- function(dlist, ordination="MDS", relPhyseq=physeq.rel, clrPhyseq=physeq.clr, csnPhyseq=physeq.CSN, nzcompPhyseq=physeq.NZcomp){
  plist <- NULL
  plist <- vector("list", 4)
  names(plist) <- c("Weighted Unifrac", "Aitchison", "Bray-Curtis", "Canberra")
  
  print("Unifrac")
  # Weighted UniFrac
  set.seed(123)
  iMDS.UniF <- ordinate(relPhyseq, ordination, distance=dlist$UniF)
  plist[[1]] <- plot_ordination(relPhyseq, iMDS.UniF, color="host_disease")
  
  print("Aitchison")
  # Aitchison
  set.seed(123)
  iMDS.Ait <- ordinate(clrPhyseq, ordination, distance=dlist$Ait)
  plist[[2]] <- plot_ordination(clrPhyseq, iMDS.Ait, color="host_disease")
  
  print("Bray")
  # Bray-Curtis
  set.seed(123)
  iMDS.Bray <- ordinate(csnPhyseq, ordination, distance=dlist$Bray)
  plist[[3]] <- plot_ordination(csnPhyseq, iMDS.Bray, color="host_disease")
  
  print("Canberra")
  # Canberra
  set.seed(123)
  iMDS.Can <- ordinate(nzcompPhyseq, ordination, distance=dlist$Can)
  plist[[4]] <- plot_ordination(nzcompPhyseq, iMDS.Can, color="host_disease")
  
  # Creating a dataframe to plot everything
  plot.df = plyr::ldply(plist, function(x) x$data)
  names(plot.df)[1] <- "distance"
  
  return(plot.df)
}
```


Now let's plot!


#### a) Fecal samples
```{r dist-run-fecal}
#________________
# FECAL DATA
#________________

# Get the distances & the plot data
dist.lopresti.fecal <- getDistances(relPhyseq = subset_samples(physeq.rel, sample_type=="stool"),
                                    clrPhyseq = subset_samples(physeq.clr, sample_type=="stool"),
                                    csnPhyseq = subset_samples(physeq.CSN, sample_type=="stool"),
                                    nzcompPhyseq = subset_samples(physeq.NZcomp, sample_type=="stool"))
plot.df.fecal <- plotDistances2D(dlist=dist.lopresti.fecal,
                                 relPhyseq = subset_samples(physeq.rel, sample_type=="stool"),
                                 clrPhyseq = subset_samples(physeq.clr, sample_type=="stool"),
                                 csnPhyseq = subset_samples(physeq.CSN, sample_type=="stool"),
                                 nzcompPhyseq = subset_samples(physeq.NZcomp, sample_type=="stool"))

# Plot
ggplot(plot.df.fecal, aes(Axis.1, Axis.2, color=host_disease))+
  geom_point(size=4, alpha=0.5)  + scale_color_manual(values = c('blue', 'red'))+
  facet_wrap(distance~., scales='free', nrow=1)+
  theme_bw()+
  theme(strip.text.x = element_text(size=20))+
  labs(color="Disease", title="Fecal samples")
# ggsave(file.path(path.plots, "distances4_MDS_stool.jpg"), height = 4, width = 15)
```

#### b) Sigmoid samples
```{r dist-run-sigmoid}
#________________
# SIGMOID DATA
#________________

# Get the distances & the plot data
dist.hugerth.sigmoid <- getDistances(relPhyseq = subset_samples(physeq.rel, sample_type=="sigmoid"),
                                   clrPhyseq = subset_samples(physeq.clr, sample_type=="sigmoid"),
                                   csnPhyseq = subset_samples(physeq.CSN, sample_type=="sigmoid"),
                                   nzcompPhyseq = subset_samples(physeq.NZcomp, sample_type=="sigmoid"))
plot.df.sigmoid <- plotDistances2D(dlist=dist.hugerth.sigmoid,
                                 relPhyseq = subset_samples(physeq.rel, sample_type=="sigmoid"),
                                 clrPhyseq = subset_samples(physeq.clr, sample_type=="sigmoid"),
                                 csnPhyseq = subset_samples(physeq.CSN, sample_type=="sigmoid"),
                                 nzcompPhyseq = subset_samples(physeq.NZcomp, sample_type=="sigmoid"))

# Plot
ggplot(plot.df.sigmoid, aes(Axis.1, Axis.2, color=host_disease))+
  geom_point(size=4, alpha=0.5)  + scale_color_manual(values = c('blue', 'red'))+
  facet_wrap(distance~., scales='free', nrow=1)+
  theme_bw()+
  theme(strip.text.x = element_text(size=20))+
  labs(color="Disease", title="Sigmoid samples")
# ggsave(file.path(path.plots, "distances4_MDS_sigmoid.jpg"), height = 4, width = 15)
```

#### c) All samples
```{r dist-run-all, fig.width = 15, fig.height = 10}
#________________
# ALL DATA
#________________

# Get the distances & the plot data
dist.hugerth <- getDistances()
plot.df <- plotDistances2D(dlist=dist.hugerth)

# Plot
p1 <- ggplot(plot.df, aes(Axis.1, Axis.2, color=host_disease))+
  geom_point(size=4, alpha=0.5)  + scale_color_manual(values = c('blue', 'red'))+
  facet_wrap(distance~., scales='free', nrow=1)+
  theme_bw()+
  theme(strip.text.x = element_text(size=20),
        axis.title.x = element_blank())+
  labs(color="Disease")
p2 <- ggplot(plot.df, aes(Axis.1, Axis.2, color=sample_type))+
  geom_line(aes(group=host_ID), color="black", lwd=0.1)+
  geom_point(size=4, alpha=0.5)  + scale_color_manual(values = c('#33CC00', 'purple'))+
  facet_wrap(distance~., scales='free', nrow=1)+
  theme_bw()+
  theme(strip.text.x = element_blank())+
  labs(color="Sample type")
ggpubr::ggarrange(p1,p2, nrow=2)
# ggsave(file.path(path.plots, "distances4_MDS_all.jpg"), height = 8, width = 15)
```




### 2. Plot in 3D

For better visualization, we will also take a glance at reduction to 3D.

```{r 3D-function, eval = TRUE, echo = TRUE}
#____________________________________________________________________________________
# Plot 3D ordination
plotDistances3D <- function(d, name_dist, physeq=physeq.hugerth){
  
  # Reset parameters
  mds.3D <- NULL
  xyz <- NULL
  fig.3D <- NULL
  
  # Reduce distance matrix to 3 dimensions
  set.seed(123)
  mds.3D <- metaMDS(d, method="MDS", k=3, trace = 0)
  xyz <- scores(mds.3D, display="sites") # pull out the (x,y,z) coordinates
  
  # Plot
  fig.3D <- plot_ly(x=xyz[,1], y=xyz[,2], z=xyz[,3], type="scatter3d", mode="markers",
                    color=sample_data(physeq)$host_disease, colors = c("blue", "red"))%>%
    layout(title = paste('MDS in 3D with', name_dist, 'distance', sep = ' '))
  
  return(fig.3D)
}

```


Now let's plot!

```{r 3D-plot, fig.width = 8, fig.height = 5}
# Fecal samples
plotDistances3D(dist.hugerth.fecal$UniF, "UniFrac", subset_samples(physeq.hugerth, sample_type=="stool"))
plotDistances3D(dist.hugerth.fecal$Ait, "Aitchison", subset_samples(physeq.hugerth, sample_type=="stool"))
plotDistances3D(dist.hugerth.fecal$Canb, "Canberra", subset_samples(physeq.hugerth, sample_type=="stool"))
plotDistances3D(dist.hugerth.fecal$Bray, "Bray-Curtis", subset_samples(physeq.hugerth, sample_type=="stool"))

# Sigmoid samples
plotDistances3D(dist.hugerth.sigmoid$UniF, "UniFrac", subset_samples(physeq.hugerth, sample_type=="sigmoid"))
plotDistances3D(dist.hugerth.sigmoid$Ait, "Aitchison", subset_samples(physeq.hugerth, sample_type=="sigmoid"))
plotDistances3D(dist.hugerth.sigmoid$Canb, "Canberra", subset_samples(physeq.hugerth, sample_type=="sigmoid"))
plotDistances3D(dist.hugerth.sigmoid$Bray, "Bray-Curtis", subset_samples(physeq.hugerth, sample_type=="sigmoid"))
```





XXXXXXXXXXXXXXXXXXXXXXX







#############################
## HIERARCHICAL CLUSTERING ##
#############################

```{r hierarchical-clust, fig.width = 10, fig.width = 7}
Heatmaps <- function(dist_list, fontsize){
  # Weighted Unifrac
  heatmp.UniF <- pheatmap(as.matrix(dist_list$UniF), 
                          clustering_distance_rows = dist_list$UniF,
                          clustering_distance_cols = dist_list$UniF,
                          fontsize = fontsize,
                          fontsize_col = fontsize-5,
                          fontsize_row = fontsize-5,
                          annotation_col = mat_col,
                          annotation_row = mat_col,
                          annotation_colors = list(group = c('Healthy' = 'blue', 'IBS' = 'red'),
                                                   sampleType = c('intestinal mucosa' = '#CC00FF', 'stool' = '#66CC00'),
                                                   subtype = c('HC'='black', 'IBS-A'='#6600FF',
                                                               'IBS-C'='#00CC66', 'IBS-D'='#FF6633')),
                          cluster_rows = T,
                          cluster_cols = T,
                          clustering_method = 'complete', #hierarchical method
                          main = 'Weighted unifrac distance')

  # Aitchison
  heatmp.Ait <- pheatmap(as.matrix(dist_list$Ait), 
                         clustering_distance_rows = dist_list$Ait,
                         clustering_distance_cols = dist_list$Ait,
                         fontsize = fontsize,
                         fontsize_col = fontsize-5,
                         fontsize_row = fontsize-5,
                         # border_color = NA,
                         cluster_rows = T,
                         cluster_cols = T,
                         clustering_method = "complete", #hierarchical method
                         annotation_col = mat_col,
                         annotation_row = mat_col,
                         annotation_colors = list(group = c('Healthy' = 'blue', 'IBS' = 'red'),
                                                   sampleType = c('intestinal mucosa' = '#CC00FF', 'stool' = '#66CC00'),
                                                   subtype = c('HC'='black', 'IBS-A'='#6600FF',
                                                               'IBS-C'='#00CC66', 'IBS-D'='#FF6633')),
                         main = "Aitchison distance")
  
    # Canberra
  heatmp.Can <- pheatmap(as.matrix(dist_list$Can), 
                         clustering_distance_rows = dist_list$Can,
                         clustering_distance_cols = dist_list$Can,
                         fontsize = fontsize,
                         fontsize_col = fontsize-5,
                         fontsize_row = fontsize-5,
                         # border_color = NA,
                         cluster_rows = T,
                         cluster_cols = T,
                         clustering_method = "complete", #hierarchical method
                         annotation_col = mat_col,
                         annotation_row = mat_col,
                         annotation_colors = list(group = c('Healthy' = 'blue', 'IBS' = 'red'),
                                                   sampleType = c('intestinal mucosa' = '#CC00FF', 'stool' = '#66CC00'),
                                                   subtype = c('HC'='black', 'IBS-A'='#6600FF',
                                                               'IBS-C'='#00CC66', 'IBS-D'='#FF6633')),
                         main = "Canberra distance")
  
  # Bray-Curtis
  heatmp.Bray <- pheatmap(as.matrix(dist_list$Bray), 
                         clustering_distance_rows = dist_list$Bray,
                         clustering_distance_cols = dist_list$Bray,
                         fontsize = fontsize,
                         fontsize_col = fontsize-5,
                         fontsize_row = fontsize-5,
                         # border_color = NA,
                         cluster_rows = T,
                         cluster_cols = T,
                         clustering_method = "complete", #hierarchical method
                         annotation_col = mat_col,
                         annotation_row = mat_col,
                         annotation_colors = list(group = c('Healthy' = 'blue', 'IBS' = 'red'),
                                                   sampleType = c('intestinal mucosa' = '#CC00FF', 'stool' = '#66CC00'),
                                                   subtype = c('HC'='black', 'IBS-A'='#6600FF',
                                                               'IBS-C'='#00CC66', 'IBS-D'='#FF6633')),
                         main = "Bray-Curtis distance")
  
  return(list("UniF" = heatmp.UniF, "Ait" = heatmp.Ait, "Can" = heatmp.Can, "Bray" = heatmp.Bray))
}


# Get the heatmaps
no_glom.heatmaps <- Heatmaps(dist_list = no_glom.dist, fontsize = 8)
```



```{r export-plots-2, eval = FALSE, echo = FALSE}
# MDS 2D all distances
jpeg("./Plots/AllDistances_MDS.jpg", width = 3000, height = 2000, res = 90)
no_glom.2D.alldist +
  scale_color_manual(values = c('blue', 'red'))+
  theme(axis.text.x = element_text(size = 20),
        legend.text = element_text(size = 50),
        axis.title = element_text(size = 50),
        title = element_text(size = 50),
        strip.text = element_text(size = 50))+
  labs(title = "LoPresti dataset (2019)")
dev.off()

jpeg("./Plots/AllDistances_MDS_sampleType.jpg", width = 3000, height = 2000, res = 90)
sampleType.2D.alldist +
  scale_color_manual(values = c('#CC00FF', '#66CC00'))+
  theme(axis.text.x = element_text(size = 20),
        legend.text = element_text(size = 50),
        axis.title = element_text(size = 50),
        title = element_text(size = 70),
        strip.text = element_text(size = 50))+
  labs(title = "LoPresti dataset (2019)")
dev.off()

jpeg("./Plots/AllDistances_MDS_subtypes.jpg", width = 3000, height = 2000, res = 90)
subtypes.2D.alldist +
  scale_color_manual(values = c('black', '#6600FF', '#00CC66', '#FF6633'))+
  theme(axis.text.x = element_text(size = 20),
        legend.text = element_text(size = 50),
        axis.title = element_text(size = 50),
        title = element_text(size = 50),
        strip.text = element_text(size = 50))
dev.off()


# Heatmaps
jpeg("./Plots/plot_heatmap_aitchison.jpg", width = 1400, height = 1000, res = 150)
no_glom.heatmaps$Ait
dev.off()

jpeg("./Plots/plot_heatmap_UniF.jpg", width = 1400, height = 1000, res = 150)
no_glom.heatmaps$UniF
dev.off()

jpeg("./Plots/plot_heatmap_canberra.jpg", width = 1400, height = 1000, res = 150)
no_glom.heatmaps$Can
dev.off()

jpeg("./Plots/plot_heatmap_bray.jpg", width = 1400, height = 1000, res = 150)
no_glom.heatmaps$Bray
dev.off()

```


################################
## REPRODUCE PLOTS FROM PAPER ##
################################

```{r reproduce-plots, fig.width = 7}
# Stool samples
stool.table <- subset_samples(physeq_best, env_material == "stool") %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt()                                             # Melt to long format

stool.table$host_disease <- factor(stool.table$host_disease, levels = c('IBS', 'Healthy')) # have same order as paper

ggplot(stool.table, aes(x = host_disease, y = Abundance, fill = Phylum))+
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(size = 8))+
  labs(x = "Samples", y = "Relative abundance", title = "Stool samples")

# Mucosa samples
mucosa.table <- subset_samples(physeq_best, env_material == "intestinal mucosa") %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt()                                             # Melt to long format

mucosa.table$host_disease <- factor(mucosa.table$host_disease, levels = c('IBS', 'Healthy')) # have same order as paper

ggplot(mucosa.table, aes(x = host_disease, y = Abundance, fill = Phylum))+
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(size = 8))+
  labs(x = "Samples", y = "Relative abundance", title = "Intestinal mucosa samples")
```


```{r export-plots-3, eval = FALSE, echo = FALSE}
jpeg("./Plots/fig1_stool.jpg", height = 1600, width = 1500, res = 300)
ggplot(stool.table, aes(x = host_disease, y = Abundance, fill = Phylum))+
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(size = 8))+
  labs(y = "Relative abundance", title = "Stool samples")
dev.off()

jpeg("./Plots/fig3_mucosa.jpg", height = 1600, width = 1500, res = 300)
ggplot(mucosa.table, aes(x = host_disease, y = Abundance, fill = Phylum))+
  geom_bar(stat = "identity", position = "fill") +
  theme(axis.text.x = element_text(size = 8))+
  labs(y = "Relative abundance", title = "Intestinal mucosa samples")
dev.off()
```





```{r}
# Function that will allow to color the leaves of the tree according to sample type (biopsy vs stool)
color_leaf<<-function(n){
    if(is.leaf(n)){
        # take the current attributes
        a=attributes(n)
        
        # deduce the line in the original data, to get the corresponding sampleType status
        line=match(attributes(n)$label, plot.df$Run)
        sampleType=plot.df[line, "sample_type"];
            if(sampleType=="sigmoid mucosa"){col_sampleType="#CC00FF"};if(sampleType=="stool"){col_sampleType="#66CC00"}
        
        #Modification of leaf attribute
        attr(n,"nodePar") <- c(a$nodePar, list(cex=1.5, lab.cex=0.6, pch=20, col=col_sampleType, lab.font=1, lab.cex=1))
        }
    return(n)
}


plot.df
```

```{r fig.height = 3}
hc.ward <- hclust(dist.thesis$Ait, method = "ward.D") #clustering 
dend.ward <- as.dendrogram(hc.ward) # Make into dendograms

dend.ward <- dendrapply(dend.ward, color_leaf) # Apply the color attributes to the dendograms

jpeg("/Users/scarcy/Desktop/Praktikum/Thesis/hclust_lopresti.jpg", height = 2000, width = 1000, res = 200)
plot(dend.ward , main="Hierarchical clustering on Aitchison distance (ward linkage)", horiz = TRUE, leaflab = "none",
     edgePar = list(lwd = 1))
axis(side = 1, lwd = 4)
legend(50, 64, 
     legend = c("Sigmoid mucosa" , "Stool"), 
     col = c("#CC00FF", "#66CC00"), 
     pch = c(20,20), bty = "n",  pt.cex = 1.5, cex = 1, 
     text.col = "black", horiz = FALSE, inset = c(0, 0.1))
dev.off()
```

