---
title: "Prediction models IBD/IBS"
author: "alexa kur"
date: '7 Mar 2017'
output: html_document
---
  
# prediction models 

## Load libraries
```{r load_libraries, error = FALSE,message=FALSE}
library(pROC)
library(glmnet)
library(knitr)
library(cvAUC)
library(caret)
library(ROCR)
```

##define prediction function

We use Lasso regression to perform model selection. 10-fold cross-validation: within each fold, 90% of the samples used to build the model, the rest 10% is used to estimate AUC.


```{r}
run_CV = function (predictors, response, non_penalized, alpha = 0.5, nfold.ex = 10, nfold.int = 5,s="lambda.min",family="binomial"){
  result = list()
  result$models = list()
  result$folds = list()
  result$predictions = list()
  result$roc = list()
  result$auc = c()
  result$predictions_zero = list()
  
  result$models_zero = list()
  result$predictions_zero = list()
  result$roc_zero = list()
  result$auc_zero = c()
  
  samples = 1:nrow(predictors)
  folds = createFolds(samples, k = nfold.ex)
  
  #model training
  for(i in names(folds)){
    #print(paste("Fold",i,"started"))
    result$folds[[i]] = folds[[i]]
    test.x=as.matrix(predictors[folds[[i]],])
    test.y=response[folds[[i]]]
    train.x=as.matrix(predictors[-folds[[i]],])
    train.y=response[-folds[[i]]]
    non_penalized.train.x = non_penalized[-folds[[i]],]
    non_penalized.test.x = non_penalized[folds[[i]],]

    #zero model
    penalty = rep(0,ncol(non_penalized.train.x))
    model.zero = glmnet(non_penalized.train.x, train.y,lambda = penalty,alpha = alpha,family = family, type.measure="class")
    result$models_zero[[i]] = model.zero
    predict.test.zero = predict(model.zero, type="class", newx=data.matrix(non_penalized.test.x))
    result$predictions_zero[[i]] = cbind(test.y,predict.test.zero[,1])
    colnames(result$predictions_zero[[i]]) = c("real","predicted")
    print(test.y)
    print(as.numeric(predict.test.zero[,1]))
    result$roc_zero[[i]] = roc(test.y, as.numeric(predict.test.zero[,1]))
    result$auc_zero[i] = result$roc_zero[[i]]$auc
    
    #work model
    #print("full model run")
    train.all.x = cbind(non_penalized.train.x,train.x)
    test.all.x = cbind(non_penalized.test.x,test.x)
    penalty = c(rep(0,ncol(non_penalized.train.x)),rep(1,ncol(train.x)))
    
    model = cv.glmnet(data.matrix(train.all.x), train.y,
                      alpha = alpha,penalty.factor = penalty,nfolds = nfold.int,family = family, type.measure="class")
    result$models[[i]] = model
    predict.test = predict(model, type="class", newx = data.matrix(test.all.x), s=s)
    result$predictions[[i]] = cbind(test.y,predict.test)
    colnames(result$predictions[[i]]) = c("real","predicted")
    result$roc[[i]] = roc(test.y, as.numeric(predict.test[,1]))
    result$auc[i] = result$roc[[i]]$auc
    
  }
    #rocr statistics
  return(result)
}


parse_results = function(pred_object){
  betas = matrix(nrow = nrow(pred_object$models$Fold01$glmnet.fit$beta),ncol = length(names(pred_object$models)),
                 dimnames = list(rownames(pred_object$models$Fold01$glmnet.fit$beta),names(pred_object$models)))
  var_orders = matrix(nrow = nrow(pred_object$models$Fold01$glmnet.fit$beta),ncol = length(names(pred_object$models)),
                 dimnames = list(rownames(pred_object$models$Fold01$glmnet.fit$beta),names(pred_object$models)))
  for(i in names(pred_object$models)){
    n = which(pred_object$models[[i]]$glmnet.fit$lambda == pred_object$models[[i]]$lambda.min)
    betas[,i] = pred_object$models[[i]]$glmnet.fit$beta[,n]
    var_orders[,i] = order(abs(betas[,i]),decreasing = F)
    }
  return(list(betas = betas,
              orders = var_orders)
  )
}
```

## Load data

Microbiome features gained from BRAKEN data. Only genera and species are inlcuded in the model. For highly correlated features (R>0.9) only one representative feature is included in the list of predictors. Feauture abundances are log-transformed. 

```{r data-preparation}
# Import Zhu
physeq.zhu
physeq.pozuelo

# Aggregate
genus.zhu <- physeq.zhu %>%
    tax_glom(taxrank = "Genus") %>%                     # agglomerate at genus level
    transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
    psmelt()                                             # Melt to long format

# Get into matrix
gmatrix.zhu <- acast(genus.zhu, Sample ~ Genus, sum, value.var = 'Abundance') # transform into matrix
```


```{r data_loading}
# TAXA
#taxonomies = read.table("taxonomy.txt",sep="\t")
all_taxa_log = log(as.matrix(gmatrix.zhu+5e-11)) #149 genera
all_taxa_log = log(data.frame(otu_table(subset_samples(physeq.pozuelo, Collection=="2nd")))+1) #21,146
#all_taxa_log = all_taxa_log[,-c(465,466,467,477)]
#all_taxa_log = all_taxa_log[,grep("g__",colnames(all_taxa_log))] # select only genus
all_taxa_log = all_taxa_log[,-findCorrelation(cor(all_taxa_log,method = "spearman"),cutoff= 0.9)] #110 genera

# OTHER PREDICTORS
#basic_predictors = as.matrix(read.table("baseModel_predictors.txt",sep="\t")) # bmi, age, ...
basic_predictors = as.data.frame(sample_data(physeq.zhu)[,c("host_sex", "host_age")]) # bmi, age, ...
basic_predictors$host_sex <- as.character(basic_predictors$host_sex)

# DISEASE
#disease = read.table("disease.txt",as.is = T,sep="\t")[,1]
disease = as.data.frame(sample_data(physeq.zhu)[,"host_disease"])
disease = as.integer(factor(disease$host_disease, levels = c("Healthy","IBS")))-1

#both = cbind(all_taxa_log,calprotectin)

#pathCal = cbind(all_pathways_log,calprotectin)
#taxaPathCal = cbind(all_pathways_log,all_taxa_log,calprotectin)


#LLD

#taxonomies_lld = read.table("taxonomy_lld.txt",header=T,row.names=1)
#taxonomies_lld = taxonomies_lld[,colnames(all_taxa_log)]

#basic_predictors_lld = as.matrix(read.table("base_lld.txt"))

#calprotectin_lld = read.table("calprotectin_lld.txt")[,1]

#disease_lld = read.table("disease_lld.txt",as.is = T)[,1]
#disease_lld = as.integer(factor(disease_lld,levels = c("S","D")))-1
```

## Feature selection

### Lasso-based prediction with 10-fold cross-validation

Base model includes age, gender and BMI without penalization.
Calprotectin model uncludes all covariates mentioned (without penalization) and calprotectin level.
Microbiome model inlcudes base predictors and microbiome feautures (described in previous block).

In the tables, 20 bacterial features with top effects are presented for two models:

1. Base predictors and microbiome features
2. Base predictors, calprotectin, microbiome features

```{r disease_type predictions,warning=F}
set.seed(2)
cv_t_result = run_CV(as.data.frame(all_taxa_log),disease,basic_predictors)

set.seed(2)
cv_t_result_scaled = run_CV(scale(all_taxa_log),disease,basic_predictors)


pp = parse_results(cv_t_result_scaled)

#set.seed(2)
#cv_tc_result = run_CV(both,disease,basic_predictors)
#set.seed(2)
#cv_p_result = run_CV(all_pathways_log,disease,basic_predictors)
#set.seed(2)
#cv_pc_result = run_CV(pathCal,disease,basic_predictors)
#set.seed(2)
#cv_tpc_result = run_CV(taxaPathCal,disease,basic_predictors)
#set.seed(2)
#cv_calprot_result = run_CV(matrix(calprotectin,ncol = 1),disease,basic_predictors)
p = parse_results(cv_t_result)
#pc = parse_results(cv_tc_result)
kable(rowMeans(p$betas[-c(1:3),])[names(head(sort(abs(rowMeans(p$betas[-c(1:3),])),decreasing = T),n = 20))])
#kable(rowMeans(pc$betas[-c(1:3),])[names(head(sort(abs(rowMeans(pc$betas[-c(1:3),])),decreasing = T),n = 21))])
predictor_names = names(head(sort(abs(rowMeans(p$betas[-c(1:3),])),decreasing = T),n = 20))

predictor_names_scaled = names(head(sort(abs(rowMeans(pp$betas[-c(1:3),])),decreasing = T),n = 20))

#predictor_names2 = names(head(sort(abs(rowMeans(pc$betas[-c(1:3),])),decreasing = T),n = 21))

```

### check goodness of fit

Bacterial features added one by one, sorted by effect size mean in 10 folds. 

```{r goodness,warning=F}
set.seed(3)
samples = 1:nrow(all_taxa_log)
folds = createFolds(samples,k = 10)
result = data.frame(features = predictor_names,aucz = rep(NA,20),aucTrain = NA,aucTest= NA)
for(j in 1:20){
  predictors = as.data.frame(all_taxa_log[,predictor_names[1:j]])
  auc = c()
  aucz = c()
  auct = c()
  for(i in names(folds)){
    #print(paste("Fold",i,"started"))
    test.x=as.matrix(predictors[folds[[i]],])
    test.y=as.factor(disease[folds[[i]]])
    train.x=as.matrix(predictors[-folds[[i]],])
    train.y=as.factor(disease[-folds[[i]]])
    non_penalized.train.x = basic_predictors[-folds[[i]],]
    non_penalized.test.x = basic_predictors[folds[[i]],]
    glm1 = glm(train.y ~ .,data = as.data.frame(cbind(non_penalized.train.x,train.x)),family = "binomial")
    pred = predict(glm1,newdata = as.data.frame(cbind(non_penalized.test.x,test.x)))
    auc = c(auc,roc(test.y,pred)$auc)
    predt = predict(glm1,newdata = as.data.frame(cbind(non_penalized.train.x,train.x)))
    auct = c(auc,roc(train.y,predt)$auc)
    glm1z = glm(train.y ~ .,data = as.data.frame(non_penalized.train.x),family = "binomial")
    predz = predict(glm1z,newdata = as.data.frame(non_penalized.test.x))
    aucz = c(aucz,roc(test.y,predz)$auc)
  }
  result$aucz[j] = mean(aucz)
  result$aucTest[j] = mean(auc)
  result$aucTrain[j] = mean(auct)
}
result_2 = result
kable(result)
```


### check goodness of fit, with calprotectin

Bacterial features added one by one, sorted by effect size mean in 10 folds.

```{r goodness_calprotectin,warning=F}
both_predictors = list()
both_labels = list()
mb_predictors = list()
mb_labels = list()
zero_predictors = list()
zero_labels = list()
cal_predictors = list()
cal_labels = list()

set.seed(3)
samples = 1:nrow(all_taxa_log)
folds = createFolds(samples,k = 10)
result = data.frame(features = predictor_names2,aucz = rep(NA,21),aucTrain = NA,aucTest= NA)
for(j in 1:21){
  predictors = as.data.frame(both[,predictor_names2[1:j]])
  auc = c()
  aucz = c()
  auct = c()
  for(i in names(folds)){
    #print(paste("Fold",i,"started"))
    test.x=as.matrix(predictors[folds[[i]],])
    test.y=as.factor(disease[folds[[i]]])
    train.x=as.matrix(predictors[-folds[[i]],])
    train.y=as.factor(disease[-folds[[i]]])
    non_penalized.train.x = basic_predictors[-folds[[i]],]
    non_penalized.test.x = basic_predictors[folds[[i]],]
    glm1 = glm(train.y ~ .,data = as.data.frame(cbind(non_penalized.train.x,train.x)),family = "binomial")
    pred = predict(glm1,newdata = as.data.frame(cbind(non_penalized.test.x,test.x)))
    auc = c(auc,roc(test.y,pred)$auc)
    predt = predict(glm1,newdata = as.data.frame(cbind(non_penalized.train.x,train.x)))
    auct = c(auc,roc(train.y,predt)$auc)
    glm1z = glm(train.y ~ .,data = as.data.frame(non_penalized.train.x),family = "binomial")
    predz = predict(glm1z,newdata = as.data.frame(non_penalized.test.x))
    aucz = c(aucz,roc(test.y,predz)$auc)
    if(j == 1){
      cal_predictors[[i]]=roc(test.y,pred)$original.predictor
      cal_labels[[i]] = roc(test.y,pred)$original.response
      zero_predictors[[i]] = roc(test.y,predz)$original.predictor
      zero_labels[[i]] = roc(test.y,predz)$original.response
    }
    if(j ==21){
      both_predictors[[i]]=roc(test.y,pred)$original.predictor
      both_labels[[i]] = roc(test.y,pred)$original.response
    }
  }
  result$aucz[j] = mean(aucz)
  result$aucTest[j] = mean(auc)
  result$aucTrain[j] = mean(auct)
}
kable(result)
```

### AUC plot

```{r fig.height=5,fig.width=5,warning=F}

# for(i in names(cv_t_result$folds)){
#   both_predictors[[i]] = cv_tc_result$roc[[i]]$original.predictor
#   both_labels[[i]] = cv_tc_result$roc[[i]]$original.response
#   mb_predictors[[i]] = cv_t_result$roc[[i]]$original.predictor
#   mb_labels[[i]] = cv_t_result$roc[[i]]$original.response
#   cal_predictors[[i]] = cv_calprot_result$roc[[i]]$original.predictor
#   cal_labels[[i]] = cv_calprot_result$roc[[i]]$original.response
#   zero_predictors[[i]] = cv_t_result$roc_zero[[i]]$original.predictor
#   zero_labels[[i]] = cv_t_result$roc_zero[[i]]$original.response}
 cvAUC_both = cvAUC(both_predictors,both_labels)
#cvAUC_mb = cvAUC(mb_predictors,mb_labels)
cvAUC_cal = cvAUC(cal_predictors,cal_labels)
cvAUC_zero = cvAUC(zero_predictors,zero_labels)

plot(cvAUC_both$perf,col = "grey82",main = "IBD/IBS prediction")
legend("bottomright",c("age + sex + calprotectin + top 20 bacterial taxa","age + sex + calprotectin","age + sex"),col = c("red","skyblue4","black"),lty=c(1,1,1),lwd = 2 )
plot(cvAUC_both$perf,col = "red",lwd =2, add=T,avg = "vertical")
plot(cvAUC_cal$perf,col = "skyblue4", lwd = 2, add=T,avg = "vertical")
plot(cvAUC_zero$perf,col = "black", lwd = 2, add=T,avg = "vertical")
```

### Formal test on comparing AUCs
```{r,warnings = F}
table_auc = data.frame(zero_pred = cv_t_result$auc_zero,
                   calprot = cv_calprot_result$auc,
                mb=cv_t_result$auc,
                 mb_calprot = cv_tc_result$auc,
                path = cv_p_result$auc,
                 pathCal = cv_pc_result$auc,
                 taxapathcal = cv_tpc_result$auc)
wilcox_test_results = foreach(i =2:ncol(table_auc),.combine = rbind)%do%{
  foreach(j=1:(i-1),.combine = rbind)%do%{
    w = wilcox.test(table_auc[,i],table_auc[,j],paired = T)
    data.frame(colnames(table_auc)[i],colnames(table_auc)[j],w$p.value)
  }
}
kable(wilcox_test_results)
```

