---
title: "R Notebook"
output: html_notebook
---

##########
# IMPORT #
##########

# Libraries
```{r import-libraries}
#install.packages("readxl")
library("readxl")
library(data.table)
library(tidyverse)
```

# Data
```{r import-data}
path <- "~/Projects/IBS_Meta-analysis_16S"

# Metadata table downloaded from the SRA
sra_metadata <- read.csv(file.path(path, "data/analysis-individual/Hugerth-2019/HugerthSraRunTable.txt"), header=TRUE, sep = ",")
#head(sra_metadata)

# Import the supplementary ASV table from the paper
paper_supp <- read_excel(file.path(path, "data/analysis-individual/Hugerth-2019/00_Metadata-Hugerth/Paper_Samples/ASV_table_PAPER.xlsx"))
head(paper_supp)

# Import the "suggested" runs from the paper (samples ERS3379832-ERS3380418)
paper_runs <- read.csv(file.path(path, "data/analysis-individual/Hugerth-2019/00_Metadata-Hugerth/paper_suggested.csv"), header=TRUE, sep = ",")
head(paper_runs)
```


##################
# CLEANUP TABLES #
##################

# SRA METADATA TABLE

The SRA metadata table was downloaded from the SRA with the accession number PRJEB31817.

```{r sra-metadata}
########
# Keep only relevant columns
col <- c("Run", "Collection_Date", "environment_.biome.", "host_body_product", "Host_Age", "host_body.mass_index", "host_subject_ID", "sample_storage_duration", "sequencing_run", "Extraction_Plate", "host_disease_status", "host_sex", "gastrointestinal_tract_disorder")

# CLEANUP SRADF
sraDF <- sra_metadata %>%
  select(col) %>%
  # remove 26 "bioreactor" samples (controls)
  filter(environment_.biome. != "bioreactor") %>% 
  select(-environment_.biome.) %>%
  # rename columns
  rename_with(tolower) %>%
  rename(Run=run,
         sample_type=host_body_product,
         host_bmi=host_body.mass_index,
         host_ID=host_subject_id,
         host_psy=host_disease_status,
         host_disease=gastrointestinal_tract_disorder) %>%
  # rename sample_type
  mutate(sample_type=replace(sample_type, sample_type=="Mucosal tissue", "sigmoid"),
         sample_type=replace(sample_type, sample_type=="Faeces", "stool")) %>%
  # rename IBS & healthy in host_disease
  mutate(host_disease=gsub("IBS.*)", "IBS", host_disease),
         host_disease=replace(host_disease, host_disease=="none", "Healthy"))
  
head(sraDF)



#######
# Create a new dataframe with only ibs/hc samples
sraDF.ibs <- sraDF %>%
  filter(host_disease == "Healthy" | host_disease == "IBS")

table(sraDF.ibs$host_disease) # sanity check ok (hc 467 - ibs 140)
```



# PAPER SUPPLEMENTARY

The supplementary material with the ASV table of the authors was downloaded directly from the paper's website https://gut.bmj.com/content/69/6/1076

```{r paper-supplementary}
# Create dataframe with the sample names (found as column names in the ASV table)
paperDF <- data.frame("raw_names" = colnames(paper_supp)[2:562])
head(paperDF)
dim(paperDF) # there are 561 rows (as there are 561 samples in the paper)

# The sample names look like "X170429_27.547_0039_sigmoideum"
# Which corresponds to "sequencing run _ extraction plate _ subject ID _ sample type"

# We need to "split" the sample names
paperDF <- paperDF %>%
  separate(col=raw_names, into=c("sequencing_run", "Extraction_Plate", "host_subject_ID", "host_body_product"), sep="_", remove=FALSE) %>%
  # rename columns
  rename_with(tolower) %>%
  rename(host_ID=host_subject_id,
         sample_type=host_body_product) %>%
  # cleanup
  mutate(sequencing_run=gsub("X", "", sequencing_run),
         extraction_plate=gsub("[.]", "-", extraction_plate),
         host_ID=as.numeric(host_ID)) %>%
  # rename sample_type
  mutate(sample_type=replace(sample_type, sample_type=="sigmoideum", "sigmoid"),
         sample_type=replace(sample_type, sample_type=="feces", "stool"))
  
head(paperDF) 
```


Now let's assign host_disease of the samples included in the paper. We have information on the sample_type, extraction_plate, sequencing_run and individual ID, which should be enough to identify the samples in the sraDF.

```{r paper-supplementary-2}
#####################################
# CROSS-REFERENCE WITH SRA METADATA #

# Add SRA columns to paperDF
sharedcol <- c("sequencing_run", "extraction_plate", "host_ID", "sample_type")
paperDF <- left_join(paperDF, sraDF, by=sharedcol)
dim(paperDF) # 562 rows, when we had 561 before...


# Check if some samples had double entry in SRA metadata
paperDF %>%
  group_by(raw_names) %>%
  filter(n()>1)
# => 1 sample has 2 entries in the SRA metadata (Healthy, age 41, bmi 25.71, collection_date 17/01/2006, but inconsistent gender)


# Check if some samples were unassigned in the SRA metadata
paperDF %>%
  filter(is.na(Run))
# => 4 samples are unassigned (not deposited on SRA database?...)


# DISEASE PHENOTYPE OF THE SAMPLES INCLUDED IN PAPER
paperDF %>%
  count(host_disease)
# take into account that one healthy sample got a double match into the SRA
# => 321 healthy samples (paper describes 466)
# => 96 IBS samples (paper describes 95)
# => rest is non-IBS and non-healthy (n=144)
```



# SAMPLES TO DOWNLOAD AS INDICATED IN PAPER

*The paper indicates to download samples ERS3379832-ERS3380418*. That's a total of 587 samples. Considering that there are 26 control samples, that would mean there are 561 IBS or healthy samples (which matches the number of samples analyzed in the paper).

Now, the _SraRunInfo.csv_ file for samples ERS3379832-ERS3380418 was downloaded on the SRA database. Let's import it, and compare the Run numbers from that file, with the Run numbers we indirectly obtained from the supplementary ASV table.

```{r suggested-runs}
# CLEANUP PAPERRUNS DF
paperRuns.DF <- paper_runs %>%
  select(Run, Sample) %>%
  # add SRA metadata to have info on samples ERS3379832-ERS3380418
  left_join(sra_metadata, by="Run") %>%
  # do same cleanup as for sraDF
  select(c(col, "Sample")) %>%
  filter(environment_.biome. != "bioreactor") %>% 
  select(-environment_.biome.) %>%
  rename_with(tolower) %>%
  rename(Run=run,
         sample_type=host_body_product,
         host_bmi=host_body.mass_index,
         host_ID=host_subject_id,
         host_psy=host_disease_status,
         host_disease=gastrointestinal_tract_disorder) %>%
  mutate(sample_type=replace(sample_type, sample_type=="Mucosal tissue", "sigmoid"),
         sample_type=replace(sample_type, sample_type=="Faeces", "stool")) %>%
  mutate(host_disease=gsub("IBS.*)", "IBS", host_disease),
         host_disease=replace(host_disease, host_disease=="none", "Healthy"))

# DISEASE PHENOTYPE OF THE SAMPLES SUGGESTED TO DOWNLOAD
paperRuns.DF %>%
  count(host_disease)

######
# Compare the samples from the supplementary ASV table, and the "suggested" samples to download
table(paperDF$Run %in% paperRuns.DF$Run)
# What are those Runs that don't match?
bind_rows(paperDF %>% filter(!Run %in% paperRuns.DF$Run),
          paperRuns.DF %>% filter(!Run %in% paperDF$Run)) %>%
  arrange(host_ID)
# 7 samples from supplementary table are not in samples suggested to download
# 6 samples suggested to download are not in supplementary table
# HOWEVER, these samples correspond to the sample individuals!! (same host_IDs...)
```

Overall, we have found the 561 samples included in the analysis by the paper. We found the sample names from (1) the supplementary ASV table and from (2) the samples ERS3379832-ERS3380418. They both correspond to the same samples ( _except for 6-7 samples, but they belong to the same individuals..._).

These 561 samples have the following gastrointestinal disorders:
- 321 healthy
- 96 IBS
- 134 other diseases (C.difficile infection, polyps, hyperplastic polyps, etc.)
- 10 unknown



XXXXXXXXXXXXXXXXXXXXXXXX REMOVE PART BELOW (JUNK) XXXXXXXXXXXXXXXXXXXXXXXX


#########################################
# Samples to quality filter through Dada2

```{r get-run-refs}
# Combine all relevant runs
# 1 - included samples in the paper (some are not ibs/hc)
authors_merged <- transform(authors_merged, Run = as.character(Run))
paper_runs <- authors_merged$Run[!is.na(authors_merged$Run)] #558

# 2 - ibs and hc samples from SRA database
mtDF_ibs_hc <- transform(mtDF_ibs_hc, Run = as.character(Run))
sra_runs <- mtDF_ibs_hc$Run #607

# Check that all ibs/hc samples from the paper are in the sra
table(na.omit(authors_merged[authors_merged$gastrointestinal_tract_disorder == "none", "Run"]) %in% sra_runs) # 321 hc (322 because one duplicate)
table(na.omit(authors_merged[authors_merged$gastrointestinal_tract_disorder == "IBS(irritable-bowel syndrome)", "Run"]) %in% sra_runs) # 96 IBS

# 3 - ibs and hc samples after modifying the contradicting infos in the SRA metadata
modified_runs <- read.csv("modified_runs.csv", header=TRUE)$Modified #585
modified_runs <- as.character(modified_runs)

# COMBINE ALL 3
all_runs <- unique(c(paper_runs, sra_runs, modified_runs))
all_runs <- all_runs[!is.na(all_runs)]

# Add "*" at the end and export
#all_runs <- paste0(all_runs, "*")
write.csv(all_runs, file = "../../list_files.csv", sep = "\t",
            row.names = FALSE)

# This file ERR3549141 could not be downloaded

mtDF_ibs_hc[mtDF_ibs_hc$Run == "ERR3549141",] # IBS
mtDF_ibs_hc[mtDF_ibs_hc$Run == "ERR3586005",] # healthy 
mtDF_ibs_hc[mtDF_ibs_hc$Run == "ERR3586042",] # healthy
mtDF_ibs_hc[mtDF_ibs_hc$Run == "ERR3586312",] # healthy
#authors_merged[authors_merged$Run == "ERR3586004",]
```


Let's build the metadata table to use in the phyloseq object for the data analysis.
```{r quick-metadata-to-delete}
# Import the metadata table with only IBS/HC that has been modified to remove inconsistencies
modif_sra <- read.csv("../../modif_metadata.csv", header=TRUE) #585
modif_sra <- transform(modif_sra, Run = as.character(Run))
modif_sra <- transform(modif_sra, host_disease = as.character(host_disease))
levels(modif_sra$sample_type) <- sub("^intestinal mucosa$", "sigmoid", levels(modif_sra$sample_type))
head(modif_sra)

# Get the samples that went through Dada2
seqtable.nochim <- readRDS("/Users/scarcy/Desktop/Praktikum/Data/Hugherth_2020/ASVtable_final.rds")
rownames(seqtable.nochim) <- gsub("_1_filt.fastq.gz","", rownames(seqtable.nochim)) # remove the .fastq.gz extension in sample names
samples_needed <- rownames(seqtable.nochim)
#samples_needed[1:10]
table(modif_sra$Run %in% samples_needed)

# Subset the ibs/hc samples
metadata_final <- modif_sra[modif_sra$Run %in% samples_needed,]
dim(metadata_final) # 582 samples

# Which samples are missing now?
table(samples_needed %in% metadata_final$Run) # 162 samples missing
samples_needed2 <- samples_needed[which(!samples_needed %in% metadata_final$Run)]
length(samples_needed2) # 162 samples missing

#######
# ADD SAMPLES NON-IBS/HC
#######
# We need to remove non-ibs/hc samples from the data analysis (we processed all paper included samples through dada2)

# 1 - For these "extra-samples", get the relevant info from SRA table
columns <- c("host_subject_ID", "Run", "host_body_product", "gastrointestinal_tract_disorder",
             "Host_Age", "host_body.mass_index", "host_sex", "host_disease_status")
DF_to_add <- sraDF[sraDF$Run %in% samples_needed2, columns]
#dim(DF_to_add) # sanity check 162 samples

# 2 - Rename columns to match our other metadata table
colnames(DF_to_add) <- c("individual", "Run", "sample_type", "host_disease",
                         "host_age", "host_bmi", "host_sex", "host_psy")
head(DF_to_add)

# 3 - Little re-formatting
DF_to_add <- transform(DF_to_add, Run = as.character(Run))
levels(DF_to_add$sample_type) <- sub("^Mucosal tissue$", "sigmoid", levels(DF_to_add$sample_type))
levels(DF_to_add$sample_type) <- sub("^Faeces$", "stool", levels(DF_to_add$sample_type))
DF_to_add$sample_type <- factor(DF_to_add$sample_type)

DF_to_add <- transform(DF_to_add, host_disease = as.character(host_disease))
DF_to_add[DF_to_add$host_disease == "none","host_disease"] <- "Healthy"
DF_to_add[DF_to_add$host_disease == "IBS (irritable-bowel syndrome)", "host_disease"] <- "IBS"
table(DF_to_add$host_disease)

head(modif_sra)
head(DF_to_add)

####### VERIFY #########
# 5 - Check that the host_ID in the "extra-samples" doesn't match any host_ID in the other metadata table
table(DF_to_add$individual %in% metadata_final$individual) # There are 62 host_IDs that can be found in the metadata table (that is only ibs/hc)
merge(x=metadata_final, y=DF_to_add, by=c("individual", "host_bmi"), all.x=FALSE) # but if we merge samples with same host_ID and same BMI, there is only 1 sample
DF_to_add[DF_to_add$individual == 5806, 'host_disease'] <- "Healthy"

# 6 - Replace host_ID in the "extra-samples"
max(modif_sra$individual)
max(DF_to_add$individual)
nbID = 6116

for (id in DF_to_add$individual){
  # If this ID exists in the ibs/hc metadata, change it
  if (id != 5806 & id %in% modif_sra$individual){
    # Give a new ID
    DF_to_add[DF_to_add$individual == id, "individual"] <- nbID
    nbID = nbID + 1 # add 1 for the next ID we will need to give
  }
}

# Sanity check
table(DF_to_add$individual %in% modif_sra$individual) # should be 1 true (one healthy sample)
table(DF_to_add$Run %in% modif_sra$Run)


#######
# MERGE IBS/HC DATAFRAME WITH PAPER SAMPLES DATAFRAME
#######

#dim(modif_sra) # 585 samples
dim(DF_to_add) # 141 samples

# Merge
metadata_final <- rbind(metadata_final, DF_to_add)
dim(metadata_final) # 744 samples
table(metadata_final$host_disease)

# Save table
write.csv(x = metadata_final, file = "../../metadata_final.csv", row.names=FALSE)
```





XXXXXXXXXXXXXXXXXXXXXXXX CLEANUP PART BELOW (PROPER REFORMATTING OF SRA METADATA) XXXXXXXXXXXXXXXXXXXXXXXX


##########################################
# REMOVE INCONSISTENCIES IN SRA METADATA #
##########################################


Let's quantify the inconsistencies

```{r metadata-inconsistencies}
# For easier read of dataframes, re-order columns
sraDF <- sraDF %>%
  # Change column order
  relocate(host_bmi, .after=Run) %>%
  relocate(host_age, .after=host_bmi) %>%
  relocate(host_sex, .after=host_age) %>%
  relocate(sample_type, .after=host_sex) %>%
  relocate(host_psy, .after=sample_type) %>%
  relocate(host_disease, .after=host_psy) %>%
  relocate(host_ID, .after=host_disease)


# Samples with same host_ID but different disease phenotype
sraDF %>%
  group_by(host_ID) %>%
  filter(n_distinct(host_disease)>1) %>%
  arrange(host_ID)
  
```



```{r filter-metadata}
#_____________________________________________
# DUPLICATE SRA METADATA
metadata <- sraDF %>%
  # create column for year-month collection date
  mutate(coldate=substr(collection_date, 0, 7))


#_____________________________________________
# STRINGENT FILTERING
stringent.df <- metadata %>%
  # group_by same BMI, age, gender, disease
  group_by(host_bmi, host_age, host_sex, host_disease, coldate) %>%
  mutate(id=group_indices()) %>%
  ungroup() %>%
  # Change column order
  relocate(coldate, .after=Run) %>%
  relocate(id, .after=host_disease) %>%
  relocate(host_ID, .after=id) %>%
  # Remove useless columns
  select(-c(coldate, host_ID)) %>%
  rename(host_ID=id) %>%
  # Keep only IBS & HC
  filter(host_disease == "IBS" | host_disease == "Healthy")


# Now that we have new identifiers, we can infer some missing data in the host_psy column
stringent.df <- stringent.df %>%
  # group by ID
  group_by(host_ID) %>%
  # replace NA values if we have several samples per individual
  fill(host_psy, .direction="downup") %>%
  # for individuals with inconsistent host_psy values, replace by NA
  mutate(host_psy=case_when(n_distinct(host_psy)>1 ~ NA_character_, TRUE ~ host_psy)) %>%
  ungroup()

```


```{r}
#_____________________________________________
# Cleanup
metadata %>%
  # group_by same BMI, age, gender, disease
  group_by(host_bmi, host_age, host_sex, host_disease) %>%
  mutate(id=group_indices()) %>%
  arrange(host_bmi, host_age, host_sex, host_disease) %>%
  # Change column order
  relocate(coldate, .after=Run) %>%
  relocate(id, .after=host_disease) %>%
  relocate(host_ID, .after=id)
  #group_vars()
# max(test$id) # 811 IDs




# Verify that individuals with same coldate, bmi, age don't have different diseases
metadata %>%
  group_by(coldate, host_bmi, host_age) %>%
  filter(n_distinct(host_disease)>1) %>%
  arrange(coldate, host_bmi, host_age) %>%
  # Change column order
  relocate(coldate, .after=Run)

# Verify individuals with same coldate, bmi, age, disease, but different genders
metadata %>%
  group_by(collection_date, host_bmi, host_age, host_disease) %>%
  filter(n_distinct(host_sex)>1) %>%
  arrange(coldate, host_bmi, host_age, host_disease) %>%
  # Change column order
  relocate(collection_date, .after=Run)
```


XXXXXXXXXXXXXXXXXXXXXXXX REMOVE PART BELOW (OLD JUNK) XXXXXXXXXXXXXXXXXXXXXXXX


Obtain the sample names that passed the quality filter with Dada2.
```{r samples-filtered}
# Get the samples that went through Dada2
seqtable.nochim <- readRDS("/Users/scarcy/Desktop/Praktikum/Data/Hugherth_2020/ASVtable_final.rds")
rownames(seqtable.nochim) <- gsub("_1_filt.fastq.gz","", rownames(seqtable.nochim)) # remove the .fastq.gz extension in sample names
samples_needed <- rownames(seqtable.nochim)

# Sanity checks
#length(samples_needed) # 744
#samples_needed[1:10]

# Get a metadata table for these samples.
```

Let's look at possible inconsistencies in the metadata from the SRA.
```{r sra-metadata-cleanup}
# To initialize before starting loop
id = 1000
#IDerr <- list("SameID_DiffBMI"=0,
#              "DiffID_SameAll"=0)


# Duplicate metadata table and sort it by increasing BMI and increasing age
new.metadata <- metadata
new.metadata <- new.metadata[order(new.metadata$host_bmi, new.metadata$host_age, new.metadata$host_disease),]

# Add collection_date column to keep only year-month
new.metadata$coldate <- new.metadata$collection_date
new.metadata$coldate <- substr(new.metadata$coldate, 0, 7)

interesting_col <- c("host_ID", "host_bmi", "host_age", "host_disease", "sample_type",
                     "host_sex", "host_psy", "collection_date")
```

```{r sra-metadata-cleanup-2}
# See before
new.metadata[,interesting_col]
mistakes.list=list("DiffPerson" = c(), "SamePerson"=c())

# For samples with the same BMI
for(bmi in bmi_range){
  # subset df
  subtable <- new.metadata[new.metadata$host_bmi == bmi,]
  
  # Give the first row a new host_ID
  subtable[1,]$host_ID <- id
  id <- id+1
  
  # If there is only 1 sample with this BMI
  #if(nrow(subtable)==1){
    # check if any other sample in metadata had the same host_ID (mistake)
    #if(new.metadata[new.metadata$host_bmi == bmi,"host_ID"] %in% new.metadata[new.metadata$host_bmi > bmi,"host_ID"]){
    #  wrong_id <- new.metadata[new.metadata$host_bmi == bmi,"host_ID"] # get the ID
    #  mistakes.list$DiffPerson <- append(mistakes.list$DiffPerson, new.metadata[new.metadata$host_ID == wrong_ID,"Run"])
    #}
  #}
  
  #______________________________________________
  # If there are 2 samples or more with this BMI
  if(nrow(subtable)>=2){
    
    #####
    # Compare row-by-row
    for (row in 2:nrow(subtable)){
      # Compare column-by-column
      for (col in c("host_bmi", "host_age", "host_disease", "host_sex", "coldate")){
        # If characteristic is the same, give same host_ID as previous row
        if (subtable[row, col] == subtable[row-1, col]){
          subtable[row,"host_ID"] <- subtable[row-1,"host_ID"]
        }
        
        # If there is only 1 characteristic  different, give new host_ID!
        else if (subtable[row, col] != subtable[row-1, col]){
          subtable[row,"host_ID"] <- id
          id <- id+1
          # Go to next row/sample
          break
        }
      }
    }
    #####
    
    #####
    # Now let's look at host_psy
    for (ID in sort(unique(subtable$host_ID))){
      
      # If there are more than 1 sample from same individual
      if(nrow(subtable[subtable$host_ID == ID,]) >= 2){
        psy.values <- subtable[subtable$host_ID == ID,"host_psy"]
        
        # If all values are NA
        if(all(is.na(psy.values))){
          subtable[subtable$host_ID == ID,"host_psy"] <- "NA"
        }
        
        # If all non-NA values agree
        else if(length(unique(psy.values[!is.na(psy.values)])) == 1){
          # replace NAs by the known value
          subtable[subtable$host_ID == ID & is.na(subtable$host_psy),"host_psy"] <- unique(psy.values[!is.na(psy.values)])
        }
        
        # If values are different
        else if(length(unique(psy.values[!is.na(psy.values)])) > 1){
          # replace host_psy by NAs
          subtable[subtable$host_ID == ID, "host_psy"] <- "NA"
        }
      }
    }
    #####
  }  
  #______________________________________________
  
  # Replace the new.metadata subset by subtable
  new.metadata[new.metadata$host_bmi == bmi,] <- subtable

}


# See after
new.metadata[,interesting_col]

```


```{r}
write.table(new.metadata, "../../modif_metadata(R).csv")

```

















